---
title: "Network Analysis of CRAN Package Dependencies"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_caption: yes
    number_sections: yes
    theme: spacelab #cerulean
    toc: yes
    toc_depth: 3
    toc_float:
      smooth_scroll: FALSE
  pdf_document: default
---

```{r knit_opts, include = FALSE}
rm(list = ls())

library(conflicted)
library(tidyverse)
library(scales)
library(purrr)
library(cowplot)
library(glue)

conflict_prefer('select',  'dplyr')
conflict_prefer('filter',  'dplyr')
conflict_prefer('lag',     'dplyr')


options(
  width = 80L,
  warn  = 1
  )

knitr::opts_chunk$set(
  tidy        = FALSE,
  cache       = FALSE,
  warning     = FALSE,
  message     = FALSE,
  fig.height  =     7,
  fig.width   =    11
  )


set.seed(42)
```

# Construct Dependency Graph

## Load CRAN Package Data

```{r retrieve_package_data, echo=TRUE, eval=TRUE}
file_name <- Sys.Date() %>% format('%Y%m%d') %>% paste0('cran_', ., '_deps.rds')

available.packages() %>%
  as_tibble() %>%
  write_rds(file = glue("data/{file_name}"))
```

We load the data into a dependency table.

```{r load_package_data, echo=TRUE}
cran_pkg_tbl <- glue("data/{file_name}") %>% read_rds()

cran_pkg_tbl %>% glimpse()
```


## Construct Depends Graph

We now construct the graph of all packages looking solely at the `Depends`
packages for now. Packages are provided in a comma-separated string, so some
parsing is required.

```{r parse_depends_fields, echo=TRUE}
construct_edges <- function(pkg_tbl, pkg_col) {
  pkg_col <- enquo(pkg_col)
  
  edges_tbl <- pkg_tbl %>%
    mutate(
      pkgs = str_split(!! pkg_col, ', *'),
      deps = map(pkgs, enframe, name = NULL)
      ) %>%
    select(Package, deps) %>%
    unnest(deps) %>%
    mutate(
      pkg_depends = gsub(' *\\(.*\\)', '', value) %>% gsub('\\n', '', .),
      pkg_depends = if_else(pkg_depends == '', NA_character_, pkg_depends),
      type        = quo_name(pkg_col)
      ) %>%
    filter(
      pkg_depends != 'R', !is.na(pkg_depends)
      )

  return(edges_tbl)
}

depends_tbl <- construct_edges(cran_pkg_tbl, Depends)
imports_tbl <- construct_edges(cran_pkg_tbl, Imports)

edgelist_tbl <- list(
    Depends = depends_tbl,
    Imports = imports_tbl
    ) %>%
  bind_rows()
```

An alternative way of doing this without the use of meta-programming ideas from
`tidyeval` is to use `tidyr` instead.


```{r construct_edgelist, echo=TRUE}
node_tbl <- cran_pkg_tbl %>%
  select(Package, NeedsCompilation)

edge_tbl <- cran_pkg_tbl %>%
  select(Package, Depends, Imports, Enhances, Suggests) %>%
  pivot_longer(
    !Package,
    names_to  = "type",
    values_to = "entry"
    ) %>%
  mutate(
    pkgs = str_split(entry, ', *'),
    deps = map(pkgs, enframe)
    ) %>%
  unnest(deps) %>%
  mutate(
    pkg_depends = gsub(' *\\(.*\\)', '', value) %>% gsub('\\n', '', .),
    pkg_depends = if_else(pkg_depends == '', NA_character_, pkg_depends)
    ) %>%
  filter(
    pkg_depends != 'R', !is.na(pkg_depends)
    ) %>%
  select(type, Package, pkg_depends) %>%
  arrange(type, Package, pkg_depends)
    

node_tbl %>% glimpse
edge_tbl %>% glimpse
```

